---
title: "Exploring Gapminder data"
author: " "
date: "Week 1: Statistical Inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  fig.height = .8 * 5,
  fig.width = .8 * 7,
  tidy.opts=list(width.cutoff = 120),  # For code
  options(width = 120)                 # for output
)
unlist(knitr::opts_chunk$get(c("fig.height", "fig.width")))

```

___Tech note___: Here is a way to make a script more portable by testing whether a package is available before installing it.




```{r packages}
if(!require(ggplot2)){install.packages("ggplot2")}
library(ggplot2)
if(!require(dplyr)){install.packages("dplyr")}
library(dplyr)
if(!require(scales)){install.packages("scales")}
library(scales)
if(!require(gapminder)){install.packages("gapminder")}
library(gapminder)
```

What variables are available, and what are their names? str() is your friend here.

```{r}
str(gapminder)
```

___Tech note___: The gapminder data set was constructed as a tibble, a generalization of a data.frame. The print() method gives an abbreviated printout. Normally, you would use head(gapminder) for this.


```{r}
gapminder
```


```{r barplot1}
ggplot(gapminder, aes(x=continent)) + geom_bar()
```

```{r barplot2}
barplot(table(gapminder$continent))
```


```{r}
gap2007<-gapminder[gapminder$year=="2007",]
boxplot(gap2007$lifeExp~gap2007$continent)
```



