---
title: "mtcars"
author: ""
date: "Week 2: Statistical Inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars}
str(mtcars)
```

```{r}
names(mtcars)
```

In this workshop we will concentrate on using mpg (miles per gallon); cyl (number of cylinders) and hp (horsepower).

Find out about the dataset:

```{r}
?mtcars
```


Look at the first 6 rows of data using the command **head**

```{r}
head(mtcars)
```

Embedding a calculated value into sentences:

The mean mpg for cars in this dataset is `r mean(mtcars$mpg)`.

Edit the statement so that it only prints to 2 decimal places:

The mean mpg for cars in this dataset is `r round(mean(mtcars$mpg),2)`.

What other summary statistics may you wish to report within the text of a document?


Create a histogram (using geom_histogram), of miles per gallon and number of cars:

* hint: use +geom_histogram(binwidth=4) *

Try to relabel your x and y axes label and include a title!

```{r}
library(ggplot2)

ggplot(mtcars, aes(mpg)) + geom_histogram(binwidth=4)
```

Edit this to show the histogram for the number of cylinders (edit the binwidth to be equal to 1).

```{r}

ggplot(mtcars, aes(cyl)) + geom_histogram(binwidth=1)

```

Would a barplot be more appropriate?

```{r}

ggplot(mtcars,aes(mpg)) + geom_bar()
```


Plot the miles per gallon versus the horsepower of the cars in this dataset.  Add in a linear regression line to this plot.

You will need to identify x and y axes here!

```{r}
ggplot(mtcars,aes(x = hp, y = mpg)) + geom_point() + geom_smooth(method = 'lm') + theme_bw()
```

Including standard errors:

```{r}

```

Use the **cor** function to find the linear correlation between the horsepower and mpg of these cars.

```{r}
cor(mtcars$mpg,mtcars$hp)
```

How would you report this within your text?

The Pearson correlation between Horsepower and Miles per gallon of these cars is `r cor(mtcars$mpg,mtcars$hp)`