---
title: "Exploratory Data Analysis for finding errors - metadata and good data management"
author: "Paul Hewson"
date: "02/03/2021"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(RSQLite)
knitr::opts_chunk$set(echo = TRUE)
```

Metadata specification
-----------------------

## Exercise 1a

Complete this table with the information you believe needs to be stored for each data set you use

```{yaml}
Data:
  Date Downloaded:
  URL:
  Metadata URL:
  Local copy of metadata:
  Licence:
  Owner:
  etc. etc. etc. etc.
```

Put your suggestions on [Padlet](https://padlet.com/texhewson/49beb939jzd9jnrt)

## Exercise 1b

Given your specification (developed under exercise 1a) complete your specification for the Health Survey for England 2011 Teaching Data Set

```{yaml}
Data:
    Title: Health survey for England 2011 Teaching Dataset
    Date Downloaded: ..........
    Coding Manual: 7402hse_2011_teaching_dataset_user_guide.pdf
    Licence Conditions: Restricted for teaching purposes
    Data Owner: DH England 2020
```


# Demonstration



1. Install RSQLite if you haven't already `install.packages("RSQLite")`
2. download the `mtcars.db` file from Blackboard and put it in your **working folder**.

The following demonstration should just "run" if you have done both 1 and 2 (and the data file is in your working folder). First see what tables we have in the database, and what columns are in the table.


```{r rsqlite}
con <- dbConnect(RSQLite::SQLite(), "mtcars.db")
dbListTables(con)
dbListFields(con, "mtcars")
```

### Pulling the data into R

We can pull in an entire table

```{r rsqlitelisting}
dbmtcars <- dbReadTable(con, "mtcars")
dbmtcars
```

Or we can use a query to be more selective.

```{r sqlitequery}
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbmtcars <- dbFetch(res)
dbmtcars
```

### Tidying up after yourself


```{r rsqlitehousekeeping}
dbClearResult(res)
dbDisconnect(con)
```

Exercise 2.
------------

## EDA with Health Survey for England 2011 Teaching Data set



1. Assuming your demonstration above ran (you have RSQLite working fine), download `hsfe.db` the 2002 Health Survey for England teaching dataset from blackboard
2. You have to bring in  `omdiaval` and `omsysval` (diastolic and systolic blood pressure) as well as `bmival` (Body Mass Index), `Sex` and `Age`.
3. Modify the query below to bring in three social variables (the query below only brings in `econact`, you need to read the user guide and pick another two)


## Pull data into R


```{r getting_hse}
con <- dbConnect(RSQLite::SQLite(), "hsfe.db")
res <- dbSendQuery(con, "SELECT Age, Sex, omdiaval, omsysval, bmival, econact
                   FROM hse2011_teaching")
hse <- dbFetch(res)
dbClearResult(res)
dbDisconnect(con)
```

## Recode missing values as NA, and give factor labels.

```{r coding_hse}
hse$Sex[hse$Sex == -9 | hse$Sex == -8] <- NA
hse$Sex <- factor(hse$Sex, labels=c("Male", "Female"))
summary(hse$Sex)
```




4. Now that you have your data in R, you should conduct an appropriate professional EDA.


Exercise 3 Organise your folders
---------------------------------


Organise your work folders so that you have a standard structure for data, scripts and metadata.



Exercise 4 Write expectations for the Health Survey for England 2011 teaching dataset
--------------------------------------------------------------------------------------

You will need to run `install.packages("assertr")` once.

## Demonstration code

```{r assertions}
library(assertr)

mtcars %>%
    verify(has_all_names("mpg", "vs", "am", "wt")) %>%
    verify(nrow(.) > 10) %>%
    verify(mpg > 0) %>%
    insist(within_n_sds(4), mpg) %>%
    assert(in_set(0,1), am, vs) %>%
    assert_rows(num_row_NAs, within_bounds(0,2), everything()) %>%
    assert_rows(col_concat, is_uniq, mpg, am, wt) %>%
    insist_rows(maha_dist, within_n_mads(10), everything()) %>%
    group_by(cyl) %>%
    summarise(avg.mpg=mean(mpg))
```

1. Define sensible and realistic expectations for the variables of interest in the Health Survey for England 2011 teaching dataset
2. Implement these Expectations using `assertr`

nb. When writing these expectations, decide whether you want to modify the query to only import data for adults

```{r adultsonly}
con <- dbConnect(RSQLite::SQLite(), "hsfe.db")
res <- dbSendQuery(con, "SELECT Age, Sex, omdiaval, omsysval, bmival, econact
                   FROM hse2011_teaching
                   WHERE Age >= 18")
hse_adult <- dbFetch(res)
dbClearResult(res)
dbDisconnect(con)

```

or whether you prefer to bring all the rows into R and use `filter` or `subset` when you need it to select only the rows for adults.

Homework
-----------


- Visit the [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) website
- Pick an old problem and review it
- Look for the new problem (data are released on a Monday) and join in `#TidyTuesday` next week.
- And join in following weeks.


