---
title: "Week 11: summary"
author: "Practical"
date: "4 May 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE, eval=TRUE}
FBook<-read.csv("../data/dataset_Facebook_v2.csv", stringsAsFactors = FALSE)
source("../mypackages.R")
source("../helperFunctions.R")
```


```{r}
str(FBook)
colnames(FBook)
```


```{r}
FBook$Post.Weekday<-as.factor(FBook$Post.Weekday)
```

```{r}
with(FBook,table(Type,Post.Weekday))
```


From this:
- may want to have Photo as the reference level of "Type" (of post) as it is the most numerous

```{r}
FBook$Type<-relevel(as.factor(FBook$Type),ref="Photo")
```


If all in the same calendar year, then months may be treated as an ordinal variable.

Why may you want to create a new variable for this?

```{r}
FBook$Month<-factor(FBook$Post.Month,ordered = TRUE, levels=1:12)
```


You Want to model the number of Lifetime post consumers using the available information.

Create a plan and work out how to implement this in R!


Want to combine information from date variable and from hour of posting - to allow for the ordering of observations from the same date!

You will need the lubridate package for this!



```{r}

mytime<-paste0(FBook$Post.Hour,":00")
# https://lubridate.tidyverse.org/reference/with_tz.html
# https://en.wikipedia.org/wiki/List_of_tz_database_time_zones

FBook$date_time <- lubridate::dmy_hm(paste(FBook$date, mytime),tz = "Europe/Lisbon")
# puts it as Lisbon time (specified timezone)
# consider how to handle timezones! 

```


```{r}

ggplot(FBook,aes(date_time,Total.Interactions, colour=Type))+geom_point()+ylab("Total Interactions")+xlab("Date")+theme_bw()
```

